---
title: "Recordings"
author: "Peter Solymos <solymos@ualberta.ca>"
---

## Introduction

Automated recording units (ARU) are increasingly being used for auditory
surveys. There are numerous advantages for using ARUs, e.g.
recordings can be stored in perpetuity to be transcribed 
later, ARUs can be programmed to record at select times and dates over long
time periods that would be prohibitive using human observers.

Bird point counts have been traditionally done by human observers. Combining
ARU data with traditional point counts thus require an understanding of
how the ARU based counts relate to counts made by human observer in the field.
The best way to approach this question is by simultaneously sampling
by two approaches: (1) human observers doing traditional point count
by registering time and distance interval an individual bird was first detected,
and (2) record the same session at the same location by an ARU to be 
identified/transcribed later in laboratory settings.

## Prerequisites

```{r aru-libs,message=TRUE,warning=FALSE}
library(bSims)                # simulations
library(detect)               # multinomial models
library(mefa4)                # count manipulation
library(paired)               # paired sampling data
```

## Paired sampling

The expected value of the total count (single species) in a 10-minutes time interval
using human observer (subscript $H$) based unlimited radius point count may be written as:
$E[Y_{H}] = D A_{H} p_{H}$ where $Y_{H}$ is the count, $D$ is population density, $A$ is 
the area sampled, $p_{H}$ is the probability that an average individual of the species
is available for detection. The quantity $p_{H}$ can be estimated based on removal 
sampling utilizing the multiple time intervals. $A_{H}$ is often unknown, but can
be estimated using the effective detection radius: $A_{H}=\pi EDR_{H}^2$.
Human observer based EDR is estimated from distance sampling.

The ARU based survey (subscript $R$ for recorder) 
can distinguish individuals within distinct time intervals,
but assigning these individuals is not yet possible using a single ARU.
An ARU based count thus can be seen ans an unlimited radius point count
where the effective area sampled is unknown. The expected value for an
ARU based count for a given species may be written as:
$E[Y_{R}] = D A_{R} p_{R}$. $p_{R}$ can be estimated based on removal 
sampling utilizing the multiple time intervals from the ARU based survey.
The unknown sampling are can be written as $A_{R}=\pi EDR_{R}^2$. 
The problem is that ARU based EDR cannot directly be estimated from the data
because of the lack of multiple distance bands or individual based distance
information.

The advantage of simultaneous sampling by human observers (H) and ARUs (A)
is that population density ($D=D_{H}=D_{R}$) is identical by design.
Possible mechanisms for differences in availability of bird individuals for detection 
($p_{H}$ vs. $p_{R}$) can include differences in how detections
are made in the field vs. in laboratory (e.g. possibility of double checking).

Both $p_{H}$ and $p_{R}$ can be estimated from the data, and the equivalence
$p=p_{H}=p_{R}$ can be tested. So for the sake of simplicity, we assume that
human observer and ARU based $p$'s are equal. 
Dividing the expected values of the counts may be written as:

$$\frac{E[Y_{R}]}{E[Y_{H}]} = \frac{D A_{R} p}{D A_{R} p} = \frac{\pi EDR_{R}^2}{\pi EDR_{H}^2} = \frac{EDR_{R}^2}{EDR_{H}^2}$$

By substituting $EDR_{R}^2 = \Delta^2 EDR_{H}^2$ (and thus $EDR_{R} = \Delta EDR_{H}$) we get:

$$\frac{E[Y_{R}]}{E[Y_{H}]} = \frac{\Delta^2 EDR_{H}^2}{EDR_{H}^2} = \Delta^2$$

This means that dividing the mean counts from ARU and human observed counts
would give an estimate of the squared scaling constant ($\Delta^2$) describing the
relationship between the estimated $EDR_{H}$ and the unknown $EDR_{R}$.

## Paired data

Human observer surveys:

* 0-50, 50-100, >100 m distance bands,
* 0-3, 3-5, 5-10 minutes time intervals.

ARU surveys:

* unlimited distance,
* 10 minutes survey in 1-minute time intervals.

```{r}
paired$DISTANCE[paired$SurveyType == "ARU"] <- "ARU"
with(paired, ftable(SurveyType, Interval, DISTANCE))
```


Select a subset of species that we'll work with:

```{r}
xt <- as.matrix(Xtab(Count ~ PKEY + SPECIES, 
  data=paired[paired$SurveyType == "HUM",]))

SPP <- colnames(xt)
## number of >0 counts
ndis <- colSums(xt > 0)
## max count
maxd <- apply(xt, 2, max)
nmin <- 15
SPP <- SPP[ndis >= nmin & maxd > 1]
SPP <- SPP[!(SPP %in%
  c("CANG","COLO","COGO","COME","FRGU","BCFR","UNKN","RESQ",
  "CORA","AMCR","WOSP","WWCR","PISI","EVGR", "RUGR", "SACR",
  "NOFL"))]
SPP
```

## Availability

We estimated availability for human observer and ARU based counts 
using the time interval information. ARU based intervals were
collapsed to the 0-3-5-10 minutes intervals to match the human observer based
design. 

```{r}
xtdurH <- Xtab(Count ~ PKEY + Interval + SPECIES, 
  paired[paired$SurveyType == "HUM",])
xtdurH <- xtdurH[SPP]
xtdurR <- Xtab(Count ~ PKEY + Interval + SPECIES, 
  paired[paired$SurveyType == "ARU",])
xtdurR <- xtdurR[SPP]

Ddur <- matrix(c(3, 5, 10), nrow(xtdurH[[1]]), 3, byrow=TRUE)
Ddur2 <- rbind(Ddur, Ddur)

xdur <- nonDuplicated(paired, PKEY, TRUE)
xx <- xdur[rownames(xtdurR[[1]]),]
```

We estimated availability for species with at least 15 detections
in both subsets of the data (making sure that the total count for at least
some locations exceeded 1). We analyzed the human observer and ARU based data 
in a single model using survey type as a dummy variable. 
We tested if the estimate corresponding to survey type differed significantly
from 0 using 95\% confidence intervals.

The following table lists singing rates (`phi` 1/minute), probability of
singing in a 10-minutes interval (`p10`), number of detections (`n`),
and whether or not the confidence limits for the survey type estimate
($\beta_1$) contained 0 (i.e. not significant survey effect).

```{r}
mdurR <- list()
mdurH <- list()
mdurHR <- list()
mdurHR1 <- list()
for (spp in SPP) {
    yR <- as.matrix(xtdurR[[spp]])[,c("0-3 min","3-5 min","5-10 min")]
    yH <- as.matrix(xtdurH[[spp]])[,c("0-3 min","3-5 min","5-10 min")]
    yHR <- rbind(yH, yR)
    mdurR[[spp]] <- cmulti(yR | Ddur ~ 1, type = "rem")
    mdurH[[spp]] <- cmulti(yH | Ddur ~ 1, type = "rem")
    aru01 <- rep(0:1, each=nrow(yH))
    mdurHR[[spp]] <- cmulti(yHR | Ddur2 ~ 1, type = "rem")
    mdurHR1[[spp]] <- cmulti(yHR | Ddur2 ~ aru01, type = "rem")
}
cfR <- sapply(mdurR, coef)
cfH <- sapply(mdurH, coef)
cfHR <- sapply(mdurHR, coef)
cfHR1 <- t(sapply(mdurHR1, coef))
names(cfR) <- names(cfH) <- names(cfHR) <- names(cfHR1) <- SPP

phiR <- exp(cfR)
phiH <- exp(cfH)
phiHR <- exp(cfHR)

## confidence interval for survey type effect
ci <- t(sapply(mdurHR1, function(z) confint(z)[2,]))
## does CI contain 0?
table(0 %[]% ci)

plot(phiR ~ phiH, 
  ylim=c(0, max(phiH, phiR)), xlim=c(0, max(phiH, phiR)),
  pch=c(21, 19)[(0 %[]% ci) + 1],
  xlab=expression(phi[H]), ylab=expression(phi[R]),
  cex=0.5+2*phiHR)
abline(0,1)
```

```{block2, type='rmdexercise'}
**Exercise**

Which $\phi$ estimate should we use? Can we use `phiHR`? 
Isn't that cheating to double the sample size? 
Think about what we are conditioning on when estimating $\phi$, and what makes samples independent.
```


## Distance sampling

We estimate EDR from human observer based counts:

```{r}
## Data for EDR estimation
xtdis <- Xtab(Count ~ PKEY + DISTANCE + SPECIES, 
  data=paired[paired$SurveyType == "HUM",])
xtdis <- xtdis[SPP]
for (i in seq_len(length(xtdis)))
    xtdis[[i]] <- as.matrix(xtdis[[i]][,c("0-49 m", "50-100 m", ">100 m")])
head(xtdis$YRWA)

## distance radii
Ddis <- matrix(c(0.5, 1, Inf), nrow(xtdis[[1]]), 3, byrow=TRUE)
head(Ddis)

## predictors
xdis <- nonDuplicated(paired, PKEY, TRUE)
xdis <- xdis[rownames(xtdis[[1]]),]
```

Fitting distance sampling models for each species:

```{r results='hide'}
mdis <- pblapply(xtdis, function(Y) {
  cmulti(Y | Ddis ~ 1, xdis, type = "dis")
})
```

```{r}
tauH <- sapply(mdis, function(z) unname(exp(coef(z))))
edrH <- 100 * tauH
round(sort(edrH))
hist(edrH)
```

## Scaling constant

Counts are often modeled in a log-linear Poisson GLM. We used GLM to estimate
the unknown scaling constant from simultaneous (paired) surveys. The Poisson mean
for a count made at site $i$ by human observer is 
$\lambda_{i,H} = D_{i} \pi EDR_H^2 p$. $EDR_H$ and $p$ are estimated using distance 
sampling and removal sampling, respectively. Those estimates are used to
calculate a correction factor $C = \pi EDR_H^2 p$ which is used as an offset
on the log scale as $log(\lambda_{i,H}) = log(D_{i}) + log(C) = \beta_0 + log(C)$,
where $\beta_0$ is the intercept in the GLM model.

Following the arguments above, the Poisson mean for an ARU based count 
made at site $i$ is $\lambda_{i,R} = D_{i} \pi \Delta^2 EDR_H^2 p = D_{i} \Delta^2 C$.
On the log scale, this becomes
$log(\lambda_{i,R}) = log(D_{i}) + log(\Delta^2) + log(C) = \beta_0 + \beta_1 + log(C)$,
where $\beta_1$ is a contrast for ARU type surveys in the log-linear model.

We used survey type as a binary variable ($x_i$) with value 0 for human observers
and value 1 for ARUs. So the Poisson model is generalized as:
$log(\lambda_{i}) = \beta_0 + x_i \beta_1 + log(C)$. $\Delta$ can be 
calculated from $\beta_1$ as $\Delta = \sqrt{e^{\beta_i}}$.

We used the Poisson GLM model describe before to estimate the $\beta_1$
coefficient corresponding to survey type as binary predictor variable,
and an offset term incorporating human observer based effective area 
sampled and availability.

```{r}
phi <- phiHR
tau <- tauH

Y <- as.matrix(Xtab(Count ~ PKEYm + SPECIES, paired))
X <- nonDuplicated(paired, PKEYm, TRUE)
X <- X[rownames(Y),]
X$distur <- ifelse(X$Disturbance != "Undisturbed", 1, 0)
X$SurveyType <- relevel(X$SurveyType, "HUM")

library(lme4)
mods <- list()
aictab <- list()
Delta <- matrix(NA, length(SPP), 3)
dimnames(Delta) <- list(SPP, c("est", "lcl", "ucl"))

#spp <- "ALFL"
for (spp in SPP) {
  y <- Y[,spp]
  C <- tau[spp]^2 * pi * (1-exp(-phi[spp]))
  off <- rep(log(C), nrow(X))

  mod0 <- glm(y ~ 1, X, offset=off, family=poisson)
  mod1 <- glm(y ~ SurveyType, X, offset=off, family=poisson)
  mod2 <- glm(y ~ SurveyType + distur, X, offset=off, family=poisson)

  aic <- AIC(mod0, mod1, mod2)
  aic$delta_AIC <- aic$AIC - min(aic$AIC)
  aictab[[spp]] <- aic

  Best <- get(rownames(aic)[aic$delta_AIC == 0])
  #summary(Best)

  mods[[spp]] <- Best
  
  ## this is Monte Carlo based CI, no need for Delta method
  bb <- MASS::mvrnorm(10^4, coef(mod1), vcov(mod1))
  Delta[spp,] <- c(sqrt(exp(coef(mod1)["SurveyTypeARU"])),
    quantile(sqrt(exp(bb[,"SurveyTypeARU"])), c(0.025, 0.975)))
}

aic_support <- t(sapply(aictab, function(z) z[,3]))
round(aic_support)
```

The following table show the estimate of $\Delta$ for each species,
and the corresponding estimates of effective detection radius (EDR) in meters
and effective area sampled ($A$) in ha:

```{r,echo=FALSE,comment=NA}
Delta_summary <- data.frame(
  Species=SPP, 
  EDR_H=round(edrH),
  EDR_R=round(edrH * Delta[,"est"]),
  A_H=round(tauH^2*pi, 2), 
  A_R=round((Delta[,"est"]*tauH)^2*pi, 2),
  Delta=round(Delta, 3))
Delta_summary
```

But wait, if we started from expected values, shouldn't ratio of
the mean counts give us $\Delta^2$?
Let's see if we can get a similar $\Delta$ value from mean counts:

```{r}
(gm <- groupMeans(Y[,SPP], 1, X$SurveyType))
Delta_summary$Delta.emp <- sqrt(gm["ARU",] / gm["HUM",])

plot(Delta.est ~ Delta.emp, Delta_summary,
  col=c(2,1)[(1 %[]% Delta[,-1]) + 1])
abline(0, 1)
abline(h=1, v=1, lty=2)
```

It looks like the fancy modeling was all for nothing, 
theory prevailed. But it is always nice when things work out
as expected.

We can also see that $\Delta$ (especially the significant ones)
tended to be less than 1, indicating that overall EDR for ARUs
is slightly smaller that for human point counts. $\Delta$
was significantly different from 1 only for relatively few species.

```{block2, type='rmdexercise'}
**Exercise**

Can we pool all species' data together to estimate an overall $\Delta$
value? Would that characterize this particular ARU type well enough?
What are some of the arguments against this pooling? What might be
driving the variation across species?
```

## Data integration

Now we will pretend that we have no paired design. See how well
fixed effects can handle the data integration without calibration.

```{r}
i <- sample(levels(X$PKEY), floor(nlevels(X$PKEY)/2))
ss <- c(which(X$PKEY %in% i), which(!(X$PKEY %in% i)))

mods2 <- list()
for (spp in SPP) {
  y <- Y[ss,spp]
  C <- tau[spp]^2 * pi * (1-exp(-phi[spp]))
  off <- rep(log(C), length(ss))

  mod <- glm(y ~ SurveyType, X[ss,], offset=off, family=poisson)

  mods2[[spp]] <- mod
}
Delta_summary$Delta.fix <- sapply(mods2, function(z) {
  sqrt(exp(coef(z)[2]))
})
plot(Delta.fix ~ Delta.emp, Delta_summary)
abline(0, 1)
abline(h=1, v=1, lty=2)
```

## Exercise

Use the script below to push the fixed effects method to the limit and
see where it fails. We will explore the following two situations:
(1) sample size and number of detections is small, (2) sampling is
biased with respect to habitat strata.



```{r eval=FALSE}
X$open <- ifelse(X$Class_Name %in% c("Open Herb/Grass",
    "Open coniferous","Open Mature Deciduous","Open Mixed",
    "Open Northern","Open Young Deciduous",
    "Open Young Mixed","Poorly Drained"), 1, 0)
## proportion of samples from ARUs (original)
prop_aru <- 0.5
## proportion of ARU samples coming from open habitats
prop_open <- 0.6

n_aru <- round(nrow(X) * prop_aru)
n_hum <- nrow(X) - n_aru
w_aru <- prop_open*X$open + (1-prop_open)*(1-X$open)
w_hum <- (1-prop_open)*X$open + prop_open*(1-X$open)

id_aru <- sample(which(X$SurveyType == "ARU"), n_aru, 
  replace=TRUE, prob=w_aru[X$SurveyType == "ARU"])
id_hum <- sample(which(X$SurveyType == "HUM"), n_hum, 
  replace=TRUE, prob=w_hum[X$SurveyType == "HUM"])

ss <- c(id_aru, id_hum)
addmargins(with(X[ss,], table(open, SurveyType)))

mods3 <- list()
for (spp in SPP) {
  y <- Y[ss,spp]
  C <- tau[spp]^2 * pi * (1-exp(-phi[spp]))
  off <- rep(log(C), length(ss))

  mod <- glm(y ~ SurveyType, X[ss,], offset=off, family=poisson)

  mods3[[spp]] <- mod
}


Est <- sapply(mods3, function(z) sqrt(exp(coef(z)[2])))

plot(Est ~ Delta.emp, Delta_summary)
abline(0, 1)
abline(h=1, v=1, lty=2)
abline(lm(Est ~ Delta.emp, Delta_summary), col=2)
```

`prop_open` 0 vs. 1 leads to different deviation from the 1:1 line, eplain why.
